---
title: "Data Run"
output:
  html_document:
    df_print: paged
---

#### Description
This document assumes you have readied your expression data for an EVE analysis, using [this script](dataClean.Rmd). Below, we will run EVE on the cleaned data, and then extract important information, such as the LRTs and beta estimates, and calculate tissue specificity.

#### Instructions

```{r include=FALSE}
knitr::knit_child("dataClean.Rmd")
library(dplyr)
library(evemodel)
library(tidyr)
```

List variable names that hold the tissue datasets
```{r}
tissues <- c("brain", "gills", "liver", "spleen")
datasets <- setNames(c(paste(tissues, "_single", sep = ""), paste(tissues, "_dup", sep = "")), 
                     c(rep(c("single", "dup"), each = 4)))
```

Run the betaShared test and store results (this could take a couple of hours)
```{r eval=FALSE}
for(i in 1:length(datasets)) {
  d <- datasets[[i]] #dataset name
  d.type <- names(datasets[i]) #duplication type
  tree <- paste(d.type, "tree", sep = "_") #tree name based on duplication type
  tree %<>% get() #retrieve tree value
  data <- get(d) #retrieve dataset value
  res.var <- paste("res", d, sep = ".") #create variable name to store results to
  print(paste("Analyzing", d))
  res.val <- betaSharedTest(tree, data, colSpecies = colnames(data), cores = 4) #run betaShared test
  assign(res.var, res.val) #assign results to the variable name
}
```

Extract LRT & beta_i values
```{r}
for(i in 1:length(datasets)){
  d <- datasets[[i]]
  res <- paste("res", d, sep = ".")
  # LRTs
  LRT.var <- paste("LRT", d, sep = ".")
  LRT.val <- get(res)$LRT
  names(LRT.val) <- rownames(get(d))
  assign(LRT.var, LRT.val)
  # beta_i
  beta.var <- paste("beta", d, sep = ".")
  beta.val <- get(res)$indivBetaRes$par[,4]
  names(beta.val) <- rownames(get(d))
  assign(beta.var, beta.val)
}
```

Rearranging data for visualization
```{r}
# Combining LRTs by duplication type
LRT.single <- as.data.frame(cbind(brain = LRT.brain_single, 
                                  gills = LRT.gills_single[names(LRT.brain_single)],
                                  liver = LRT.liver_single[names(LRT.brain_single)],
                                  spleen = LRT.spleen_single[names(LRT.brain_single)]))
LRT.dup <- as.data.frame(cbind(brain = LRT.brain_dup, 
                                  gills = LRT.gills_dup[names(LRT.brain_dup)],
                                  liver = LRT.liver_dup[names(LRT.brain_dup)],
                                  spleen = LRT.spleen_dup[names(LRT.brain_dup)]))
#long form
LRT.single_long <- LRT.single %>% 
  mutate(gene = rownames(LRT.single)) %>% 
  pivot_longer(-gene, names_to = "tissue", values_to = "LRT")
LRT.dup_long <- LRT.dup %>% 
  mutate(gene = rownames(LRT.dup)) %>% 
  pivot_longer(-gene, names_to = "tissue", values_to = "LRT")
```

```{r}
# Combining beta_i by duplication type
beta.single <- as.data.frame(cbind(brain = beta.brain_single, 
                                  gills = beta.gills_single[names(beta.brain_single)],
                                  liver = beta.liver_single[names(beta.brain_single)],
                                  spleen = beta.spleen_single[names(beta.brain_single)]))
beta.shared_single <- data.frame(tissue = c("brain", "gills", "liver", "spleen"), 
                                 beta_shared = c(res.brain_single$sharedBeta, res.gills_single$sharedBeta, 
                                                 res.liver_single$sharedBeta, res.spleen_single$sharedBeta))
beta.dup <- as.data.frame(cbind(brain = beta.brain_dup, 
                               gills = beta.gills_dup[names(beta.brain_dup)],
                               liver = beta.liver_dup[names(beta.brain_dup)],
                               spleen = beta.spleen_dup[names(beta.brain_dup)]))
beta.shared_dup <- data.frame(tissue = c("brain", "gills", "liver", "spleen"), 
                                 beta_shared = c(res.brain_dup$sharedBeta, res.gills_dup$sharedBeta, 
                                                 res.liver_dup$sharedBeta, res.spleen_dup$sharedBeta))
#long form
beta.single_long <- beta.single %>% mutate(gene = rownames(beta.single)) %>% 
  pivot_longer(-gene, names_to = "tissue", values_to = "beta_i") %>%
  inner_join(., beta.shared_single, by = "tissue")
beta.dup_long <- beta.dup %>% mutate(gene = rownames(beta.dup)) %>% 
  pivot_longer(-gene, names_to = "tissue", values_to = "beta_i") %>%
  inner_join(., beta.shared_dup, by = "tissue")
```

```{r}
# Comprehensive combined dataset that includes LRTs, beta_i, beta_shared, by duplication type
combined.single <- inner_join(beta.single_long, LRT.single_long, by = c("gene", "tissue"))
combined.dup <- inner_join(beta.dup_long, LRT.dup_long, by = c("gene", "tissue"))
```

Calculate tissue specificities using [this file](tissSpec.Rmd)
```{r include=FALSE}
knitr::knit_child("tissSpec.Rmd")
```

Add tissue specificity to comprehensive datasets
```{r}
combined.single <- left_join(combined.single, tauExp.filt.long, by = c("gene", "tissue"))
combined.dup <- left_join(combined.dup, tauExp.filt.long, by = c("gene", "tissue"))
combined.single$tiss.spec[is.na(combined.single$tiss.spec)] <- "None"
combined.dup$tiss.spec[is.na(combined.dup$tiss.spec)] <- "None"
combined.single$tiss.spec <- factor(combined.single$tiss.spec, levels = c("None", "brain", "gills", "liver", "spleen", "Other"))
combined.dup$tiss.spec <- factor(combined.dup$tiss.spec, levels = c("None", "brain", "gills", "liver", "spleen", "Other"))
head(combined.single)
head(combined.dup)
```

Next step is to [visualize the data](dataVis.Rmd)
